---
title: "Post-seeding assembly processes shape seed banks more than initial restoration design"
date: "2026-06-22"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7,
  dpi = 96
)
# Note: set error = TRUE above if you want knitting to continue past a failing chunk.
```

## install packages

```{r install-packages-0}
# install.packages("here")        # Run this once in the console
# install.packages("lmerTest")
# install.packages("ggeffects")
# install.packages("tidyverse")
# install.packages("emmeans")
# install.packages("multcomp")
# install.packages("multcompView")
# install.packages("devtools")
# install.packages("vegan")
# install.packages("ggplot2")
# install.packages("ggordiplots")
# install.packages("ggpubr")
# install.packages("tidytext")
```

## load libraries

```{r load-libraries-1}
library(here)
library(tidyverse)
library(lme4)
library(lmerTest)
library(emmeans)
library(multcomp)
library(multcompView)
library(vegan)
library(ggordiplots)
library(ggpubr)
library(ggplot2)
library(car)
# pak::pak("pmartinezarbizu/pairwiseAdonis/pairwiseAdonis")
library(pairwiseAdonis)

# Set global font to Arial
theme_set(theme_gray(base_family = "Arial"))
```

## global settings and helper functions

```{r global-settings-and-helper-functions-2}
# Global factor levels
trt_levels <- c("R", "DT", "FD", "IR")
trt_labels <- c("Random", "Drought Tolerant", "Functionally Diverse", "Invasion Resistant")
# two-word labels wrapped onto two lines (second word below first) for narrow panels
trt_labels_wrap <- c("Random", "Drought\nTolerant", "Functionally\nDiverse", "Invasion\nResistant")
water_levels <- c("125", "50")
water_labels <- c("Precipitation Addition", "Precipitation Reduction")

# Shared themes
theme_custom <- function() {
  theme_minimal(base_family = "Arial") +
    theme(
      axis.text.x = element_text(size = 16, family = "Arial"),
      axis.text.y = element_text(size = 16, family = "Arial"),
      axis.title.x = element_text(size = 18, face = "bold", family = "Arial"),
      axis.title.y = element_text(size = 18, face = "bold", family = "Arial"),
      plot.title = element_text(face = "bold", size = 20, hjust = 0.5, family = "Arial"),
      legend.text = element_text(size = 14, family = "Arial"),
      legend.title = element_text(size = 16, face = "bold", family = "Arial"),
      legend.position = "bottom",
      legend.key.size = unit(1, "cm")
    )
}

theme_nmds <- function() {
  theme_minimal(base_family = "Arial") +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 16, family = "Arial"),
      axis.title = element_text(face = "bold", size = 14, family = "Arial"),
      axis.text = element_text(size = 13, family = "Arial"),
      legend.title = element_text(face = "bold", size = 14, family = "Arial"),
      legend.text = element_text(size = 13, family = "Arial"),
      legend.position = "right",
      panel.border = element_rect(linewidth = 1.5, fill = NA)
    )
}

# apply standard factor levels:
apply_factors <- function(df, trt_col = "treatment", water_col = "water") {
  if (!is.null(trt_col) && trt_col %in% names(df)) {
    df[[trt_col]] <- factor(toupper(df[[trt_col]]), levels = trt_levels)
  }
  if (!is.null(water_col) && water_col %in% names(df)) {
    df[[water_col]] <- factor(df[[water_col]], levels = water_levels)
  }
  df
}
# helper: calculate whisker positions for letter placement
calc_whisker_positions <- function(data, y_var, group_vars) {
  data %>%
    group_by(across(all_of(group_vars))) %>%
    summarise(
      Q3 = quantile(.data[[y_var]], 0.75, na.rm = TRUE),
      Q1 = quantile(.data[[y_var]], 0.25, na.rm = TRUE),
      IQR = Q3 - Q1,
      upper_whisker = min(max(.data[[y_var]], na.rm = TRUE), Q3 + 1.5 * IQR),
      .groups = "drop"
    ) %>%
    mutate(yposition = upper_whisker + 0.05)
}
# helper: Clean and sort matrix row names
clean_sort_matrix <- function(matrix_obj) {
  rownames(matrix_obj) <- gsub("-", " ", rownames(matrix_obj))
  rownames(matrix_obj) <- tolower(rownames(matrix_obj))
  rownames(matrix_obj) <- gsub("([0-9]+) (\\w+)", "\\2 \\1", rownames(matrix_obj))
  
  row_parts <- strsplit(rownames(matrix_obj), " ")
  row_sort_key <- data.frame(
    alpha = sapply(row_parts, `[`, 1),
    num = as.numeric(sapply(row_parts, `[`, 2))
  )
  sorted_rows <- order(row_sort_key$alpha, row_sort_key$num)
  matrix_obj[sorted_rows, ]
}

# helper: get CLD letters for mixed models
get_cld_letters <- function(model_obj, data_df, y_var) {
  emm <- emmeans(model_obj, ~ treatment * type * water)
  cld_df <- cld(emm, Letters = letters, reversed = TRUE, alpha = 0.05)
  cld_df$Letters <- trimws(cld_df$.group)
  cld_df <- cld_df %>% mutate(treatment = factor(treatment, levels = trt_levels))
  
  data_df %>%
    filter(!is.na(water), !is.na(type), !is.na(treatment)) %>%
    group_by(treatment, type, water) %>%
    summarise(
      Q3 = quantile(.data[[y_var]], 0.75, na.rm = TRUE),
      Q1 = quantile(.data[[y_var]], 0.25, na.rm = TRUE),
      IQR = Q3 - Q1,
      upper_whisker = min(max(.data[[y_var]], na.rm = TRUE), Q3 + 1.5 * IQR),
      .groups = "drop"
    ) %>%
    mutate(yposition = upper_whisker + 0.05) %>%
    left_join(cld_df, by = c("treatment", "type", "water"))
}
```

## load and prepare data

```{r load-and-prepare-data}
plot_metadata <- read.csv(here("plot_metadata.csv"))
species_counts <- read.csv(here("species_counts.csv"))
# View(plot_metadata)
# View(species_counts)

# unpack univariate tracking metrics
native   <- plot_metadata %>% dplyr::select(plot, treatment, water, structure, total_native)
invasive <- plot_metadata %>% dplyr::select(plot, treatment, water, structure, total_invasive)
native.rev <- plot_metadata %>% dplyr::select(plot, treatment, water, structure)

# reconstruct metadata lookup mapping for environmental factors
water.levels <- plot_metadata %>%
  dplyr::select(plot, treatment, water) %>%
  mutate(match_id = paste0(tolower(treatment), " ", plot)) %>%
  distinct()
rownames(water.levels) <- water.levels$match_id

abvg_water <- plot_metadata %>% dplyr::select(plot, treatment, water, structure)

# unpack Species Trait Matrix
seed_data <- species_counts %>%
  distinct(species_code, seed.mass, mean_abundance) %>%
  rename(code = species_code) %>%
  filter(!is.na(code))

# pivot long matrices back to wide structures for Vegan/NMDS steps
year0 <- species_counts %>%
  filter(sample_type == "year0") %>%
  dplyr::select(sample_id, species_code, abundance) %>%
  pivot_wider(names_from = species_code, values_from = abundance, values_fill = 0) %>%
  as.data.frame()
colnames(year0)[1] <- "X"

native_mat.rev <- species_counts %>%
  filter(sample_type == "year4_below") %>%
  dplyr::select(sample_id, species_code, abundance) %>%
  pivot_wider(names_from = species_code, values_from = abundance, values_fill = 0) %>%
  as.data.frame()
colnames(native_mat.rev)[1] <- "X"

abvg_4 <- species_counts %>%
  filter(sample_type == "year4_above") %>%
  dplyr::select(sample_id, species_code, abundance) %>%
  pivot_wider(names_from = species_code, values_from = abundance, values_fill = 0) %>%
  as.data.frame()
colnames(abvg_4)[1] <- "X"

# convert layout indices to factor designations
for (df_name in c("native", "invasive")) {
  df <- get(df_name)
  df$plot <- as.factor(df$plot)
  df$treatment <- as.factor(df$treatment)
  df$water <- as.factor(df$water)
  df$structure <- as.factor(df$structure)
  assign(df_name, df)
}
```

## analysis 1: native vs invasive counts

```{r analysis-1-native-vs-invasive-counts-4}
native <- apply_factors(native)
invasive <- apply_factors(invasive)

combined_df <- full_join(native, invasive, by = c("plot", "treatment", "water", "structure"))
combined_new <- combined_df[-27, ] #outlier removed

long_new <- combined_new %>%
  pivot_longer(cols = c(total_native, total_invasive), names_to = "type", values_to = "counts") %>%
  apply_factors()

# native glmer.nb 
native_model <- glmer.nb(total_native ~ treatment * water + (1|structure), data = native, family = poisson)
summary(native_model)
Anova(native_model) #type II chi-sq test

combined_model_nb <- glmer.nb(total_invasive ~ treatment * water + (1|structure), data = combined_new)
summary(combined_model_nb)
#anova(combined_model_nb)
Anova(combined_model_nb) #type II chi-sq

# post-hoc: emmeans on the actual GLMER models (type = "response" back-transforms to count scale)
emm_native   <- emmeans(native_model,      ~ treatment * water, type = "response")
emm_invasive <- emmeans(combined_model_nb, ~ treatment * water, type = "response")
# no significant interactions but using for CLD on ggplot

cld_native <- as.data.frame(cld(emm_native,   Letters = letters, reversed = TRUE, alpha = 0.05)) %>%
  mutate(type = "total_native",   Letters = trimws(.group))
cld_invasive <- as.data.frame(cld(emm_invasive, Letters = letters, reversed = TRUE, alpha = 0.05)) %>%
  mutate(type = "total_invasive", Letters = trimws(.group))

letters_df <- bind_rows(
  cld_native   %>% dplyr::select(treatment, water, type, Letters),
  cld_invasive %>% dplyr::select(treatment, water, type, Letters)
) %>%
  mutate(
    treatment = factor(toupper(as.character(treatment)), levels = trt_levels),
    water     = factor(as.character(water), levels = water_levels)
  )

plot_data <- long_new %>%
  calc_whisker_positions("counts", c("treatment", "water", "type")) %>%
  mutate(label_y = upper_whisker + 5) %>%
  left_join(letters_df, by = c("treatment", "water", "type"))
```

### Figure 1, count data

```{r ggplot1-count-data-5}
counts_boxplot <- ggplot(long_new, aes(x = treatment, y = counts, fill = interaction(water, type))) +
  geom_boxplot(position = position_dodge(width = 0.9), outlier.shape = NA) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "darkred",
               position = position_dodge(width = 0.9), aes(group = interaction(water, type))) +
  geom_text(data = plot_data,
            aes(x = treatment, y = label_y, label = Letters, group = interaction(water, type)),
            position = position_dodge(width = 0.9), vjust = 0, size = 7.5, show.legend = FALSE, family = "Arial") +
  scale_x_discrete(labels = trt_labels) +
  scale_y_continuous(labels = function(x) x / 0.01) +
  scale_fill_viridis_d(
    option = "viridis", begin = 0.2, end = 0.9, name = "",
    labels = c("125.total_invasive" = "Invasive (Precipitation Addition)",
               "50.total_invasive" = "Invasive (Precipitation Reduction)",
               "125.total_native" = "Native (Precipitation Addition)",
               "50.total_native" = "Native (Precipitation Reduction)")
  ) +
  labs(x = NULL, y = expression(bold("Seed Count (per m"^2*")"))) +
  theme_custom() +
  theme(
    panel.grid       = element_blank(),                                  # remove background grid
    axis.line        = element_line(colour = "black", linewidth = 0.8),  # solid axis lines
    axis.ticks       = element_line(colour = "black", linewidth = 0.8),
    legend.text      = element_text(size = 20, family = "Arial"),
    legend.title     = element_text(size = 20, family = "Arial"),
    legend.key.size  = unit(1.2, "cm"),
    axis.title.y     = element_text(size = 20, face = "bold", family = "Arial"),
    axis.title.x     = element_text(size = 16, face = "bold", family = "Arial"),
    axis.text.x      = element_text(size = 20, family = "Arial"),
    axis.text.y      = element_text(size = 20, family = "Arial")
  ) +
  # stack the 4 legend entries into a 2-row x 2-column block at the bottom
  guides(fill = guide_legend(nrow = 2, byrow = TRUE))
print(counts_boxplot)
ggsave("Figure_1.png", plot = counts_boxplot, width = 16, height = 8, dpi = 1200, units = "in", bg = "white")
```

## analysis 2: shannon diversity

```{r analysis-2-shannon-diversity-6}
# matrix prep
plotnames <- native_mat.rev[, 1]
native_mat.rev2 <- native_mat.rev[, -1]
rownames(native_mat.rev2) <- plotnames
native.rev <- native.rev[-75, ]    ## outlier

# Calculate diversity
shannon_native2 <- diversity(native_mat.rev2, index = "shannon")
native.rev$shannon_native2 <- shannon_native2

# Apply factors
native.rev$treatment <- factor(toupper(native.rev$treatment), levels = trt_levels)
native.rev$water <- factor(native.rev$water, levels = water_levels)

# Statistical analysis
native_div.model <- lmer(shannon_native2 ~ treatment * water + (1|structure), data = native.rev)
summary(native_div.model)
anova(native_div.model)

# Generate estimated marginal means and CLD
emm_div <- emmeans(native_div.model, ~ treatment * water)
summary(emm_div)
cld_div <- cld(emm_div, Letters = letters, reversed = TRUE, alpha = 0.05)

# Clean CLD data frame
cld_div <- as.data.frame(cld_div) %>%
  mutate(
    Letters = trimws(.group),
    treatment = factor(treatment, levels = trt_levels),
    water = factor(water, levels = water_levels)
  )

# Calculate whisker positions
letters_df_div <- native.rev %>%
  group_by(treatment, water) %>%
  summarise(
    Q3 = quantile(shannon_native2, 0.75, na.rm = TRUE),
    Q1 = quantile(shannon_native2, 0.25, na.rm = TRUE),
    IQR = Q3 - Q1,
    upper_whisker = min(max(shannon_native2, na.rm = TRUE), Q3 + 1.5 * IQR),
    .groups = "drop"
  ) %>%
  left_join(cld_div, by = c("treatment", "water")) %>%
  mutate(yposition = upper_whisker + 0.05)

# Plot
shannon_boxplot <- ggplot(native.rev, aes(x = treatment, y = shannon_native2, fill = water)) +
  geom_boxplot(position = position_dodge(width = 0.75), outlier.shape = NA) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, 
               position = position_dodge(width = 0.75), color = "darkred",
               aes(group = water)) +
  geom_text(data = letters_df_div,
            aes(x = treatment, y = yposition, label = Letters, group = water), 
            position = position_dodge(width = 0.75),
            vjust = 0, size = 6.5) + 
  scale_x_discrete(labels = trt_labels_wrap) +
  scale_y_continuous(expand = expansion(mult = c(0.07, 0.15))) +
  scale_fill_viridis_d(
    option = "viridis",
    begin = 0.2,
    end = 0.6,
    name = "",
    labels = water_labels
  ) +
  labs(x = NULL, y = "Native Shannon Diversity") +
  theme_custom() +
  theme(
    panel.grid      = element_blank(),                                  # remove background grid
    axis.line       = element_line(colour = "black", linewidth = 0.8),  # solid axis lines
    axis.ticks      = element_line(colour = "black", linewidth = 0.8),
    axis.title.y    = element_text(size = 20, face = "bold", family = "Arial"),
    axis.title.x    = element_text(size = 20, face = "bold", family = "Arial"),
    axis.text.x     = element_text(size = 20, family = "Arial"),
    axis.text.y     = element_text(size = 20, family = "Arial"),
    legend.text     = element_text(size = 20, family = "Arial"),
    legend.title    = element_text(size = 20, family = "Arial"),
    legend.key.size = unit(1.2, "cm")
  )
# shannon_boxplot is combined with richness_boxplot in the Figure 2 chunk below
```

## analysis 3: native species richness

```{r analysis-3-native-species-richness-7}
# 1. filter out the columns
native_mat.rev6 <- native_mat.rev[, !colnames(native_mat.rev) %in% c("LUAL", "MASA", "PEHE", "SACO")]

# 2. drop the first column
native_mat.rev7 <- native_mat.rev6[, -1]
rownames(native_mat.rev7) <- native_mat.rev6[, 1]
native_mat.rev.4 <- clean_sort_matrix(native_mat.rev7)

native_richness <- specnumber(native_mat.rev.4)
native.rev$richness <- native_richness
native.rev <- apply_factors(native.rev)

richness_model <- lmer(richness ~ treatment * water + (1|structure), data = native.rev)
summary(richness_model)
anova(richness_model)

emm_richness <- emmeans(richness_model, ~ treatment * water)
print(emm_richness)
cld_richness <- cld(emm_richness, Letters = letters, reversed = TRUE, alpha = 0.05)

cld_richness <- as.data.frame(cld_richness) %>%
  mutate(
    Letters = trimws(.group),
    treatment = factor(treatment, levels = trt_levels),
    water = factor(water, levels = water_levels)
  )
plot_letters_rich <- native.rev %>%
  calc_whisker_positions("richness", c("treatment", "water")) %>%
  left_join(cld_richness, by = c("treatment", "water"))
```

## Figure 2, richness panel

```{r ggplot3-richness-boxplot-8}
richness_boxplot <- ggplot(native.rev, aes(x = treatment, y = richness, fill = water)) +
  geom_boxplot(position = position_dodge(width = 0.75), outlier.shape = NA) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, 
               color = "darkred", position = position_dodge(width = 0.75),
               aes(group = water)) +
  geom_text(data = plot_letters_rich, 
            aes(x = treatment, y = yposition, label = Letters, group = water), 
            position = position_dodge(width = 0.75),  
            vjust = 0, size = 6.5) + 
  scale_x_discrete(labels = trt_labels_wrap) +
  scale_y_continuous(expand = expansion(mult = c(0.07, 0.1))) +
  scale_fill_viridis_d(
    option = "viridis",
    begin = 0.2,
    end = 0.6,
    name = "",
    labels = water_labels
  ) +
  labs(x = NULL, y = "Native Species Richness") +
  theme_custom() +
  theme(
    panel.grid      = element_blank(),                                  # remove background grid
    axis.line       = element_line(colour = "black", linewidth = 0.8),  # solid axis lines
    axis.ticks      = element_line(colour = "black", linewidth = 0.8),
    axis.title.y    = element_text(size = 20, face = "bold", family = "Arial"),
    axis.title.x    = element_text(size = 20, face = "bold", family = "Arial"),
    axis.text.x     = element_text(size = 20, family = "Arial"),
    axis.text.y     = element_text(size = 20, family = "Arial"),
    legend.text     = element_text(size = 20, family = "Arial"),
    legend.title    = element_text(size = 20, family = "Arial"),
    legend.key.size = unit(1.2, "cm")
  )
# richness_boxplot is combined with shannon_boxplot in the Figure 2 chunk below
```

## Figure 2: Shannon diversity & richness (combined)

```{r figure-2-shannon-richness-combined}
# Panel A = native Shannon diversity, Panel B = native species richness.
# Both panels share the same precipitation legend, so it is drawn once at the bottom.
figure_2 <- ggarrange(
  shannon_boxplot, richness_boxplot,
  labels        = c("A", "B"),
  ncol          = 1, nrow = 2,
  common.legend = TRUE,
  legend        = "bottom",
  font.label    = list(size = 22, face = "bold", family = "Arial")
)
print(figure_2)
ggsave("Figure_2.png", plot = figure_2, width = 10, height = 12, dpi = 1200, units = "in", bg = "white")
```

## analysis4: Pielou's J

```{r analysis4-pielou-s-j-9}
richness_vals <- specnumber(native_mat.rev.4)
shannon_vals <- diversity(native_mat.rev.4, index = "shannon")

J <- shannon_vals / log(richness_vals)

pielou_data <- data.frame(
  Sample = names(J),
  Evenness = as.numeric(J),
  stringsAsFactors = FALSE
) %>%
  separate(col = Sample, into = c("treatment", "plot"), sep = " ", remove = FALSE, convert = FALSE, fill = "right") %>%
  filter(!is.na(Evenness), !is.infinite(Evenness)) %>%
  mutate(
    treatment = toupper(treatment),
    treatment = factor(treatment, levels = trt_levels),
    plot = as.numeric(plot)
  )

print(head(pielou_data))
print(paste("Number of observations:", nrow(pielou_data)))

anova_pielou <- aov(Evenness ~ treatment, data = pielou_data)
tukey_pielou <- TukeyHSD(anova_pielou)
```

### ggplot4: pielou_boxplot

```{r ggplot4-pielou-boxplot-10}
  pielou_boxplot <- ggplot(pielou_data, aes(x = treatment, y = Evenness, fill = treatment)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  geom_jitter(width = 0.1, alpha = 0.5, size = 2) +
  scale_x_discrete(labels = trt_labels) +
  scale_fill_viridis_d(option = "viridis", begin = 0.2, end = 0.9) +
  scale_y_continuous(limits = c(0, 1.1), expand = expansion(mult = c(0.05, 0.05))) +
  labs(x = "Seed Mixes", y = "Evenness (J')", title = "Pielou's Evenness by Treatment") +
  theme_custom() +
  theme(legend.position = "none")

print(pielou_boxplot)
ggsave("pielous.j.png", plot = pielou_boxplot,width = 10, height = 6, dpi = 1200, units = "in", bg = "white" )
cat("\n=== PIELOU'S EVENNESS ANOVA ===\n")
print(summary(anova_pielou))

cat("\n=== TUKEY HSD POST-HOC TEST ===\n")
print(tukey_pielou)
```

## analysis 5: NMDS year 0

```{r analysis-5-nmds-year-0-11}
plotnames <- year0[, 1]
year0_clean <- year0 %>% dplyr::select(-1)
rownames(year0_clean) <- plotnames

rows_to_remove <- c(
  "dt 1", "dt 10", "dt 2", "dt 22", "dt 23", "dt 24", "dt 25", "dt 26", "dt 27", "dt 28", "dt 3", "dt 40",
  "dt 41", "dt 42", "dt 45", "dt 46", "dt 47", "dt 49", "dt 50", "dt 51", "dt 52", "dt 53", "dt 7", "dt 8", "dt 9",
  "fd 1", "fd 10", "fd 11", "fd 12", "fd 2", "fd 22", "fd 23", "fd 24", "fd 25", "fd 26", "fd 27", "fd 28", "fd 29",
  "fd 3", "fd 36", "fd 4", "fd 40", "fd 41", "fd 42", "fd 45", "fd 46", "fd 47", "fd 49", "fd 50", "fd 51", "fd 52",
  "fd 53", "fd 8", "fd 9", "ir 1", "ir 10", "ir 2", "ir 22", "ir 23", "ir 24", "ir 25", "ir 26", "ir 27", "ir 28",
  "ir 29", "ir 3", "ir 36", "ir 40", "ir 41", "ir 42", "ir 43", "ir 45", "ir 46", "ir 47", "ir 50", "ir 51", "ir 52",
  "ir 53", "ir 8", "ir 9", "r 1", "r 10", "r 11", "r 2", "r 22", "r 23", "r 24", "r 25", "r 26", "r 27", "r 28",
  "r 29", "r 3", "r 32", "r 4", "r 40", "r 41", "r 42", "r 43", "r 45", "r 46", "r 47", "r 50", "r 51", "r 52",
  "r 53", "r 8", "r 9", "dt 35"
)

year0_sorted <- year0_clean[!rownames(year0_clean) %in% rows_to_remove, ] %>%
  {.[order(sapply(strsplit(rownames(.), " "), `[`, 1),
           as.numeric(sapply(strsplit(rownames(.), " "), `[`, 2))), ]}

native.rel.rev <- decostand(native_mat.rev.4, method = "total")
year_4 <- native.rel.rev[, 1:32]
common_cols <- intersect(colnames(year_4), colnames(year0_sorted))
year_0 <- year0_sorted[, common_cols]
year_0[is.na(year_0)] <- 0

year_0bc <- vegdist(as.matrix(year_0), method = "bray")
ca.nmds_yr0 <- metaMDS(year_0bc, k = 3, trymax = 100, trace = FALSE)

nmds_scores_yr0 <- as.data.frame(scores(ca.nmds_yr0, display = "sites"))

rownames_yr0 <- rownames(nmds_scores_yr0)

nmds_scores_yr0 <- nmds_scores_yr0 %>%
  mutate(
    rowname_original = rownames_yr0,
    trt = tolower(sub(" .*", "", rowname_original)),
    water = as.character(water.levels[rowname_original, "water"])
  ) %>%
  mutate(
    trt = factor(trt, levels = c("r", "dt", "fd", "ir")),
    trt = recode_factor(trt,
                        "r" = "Random",
                        "dt" = "Drought Tolerant",
                        "fd" = "Functionally Diverse",
                        "ir" = "Invasion Resistant"),
    water = factor(water, levels = water_levels, labels = water_labels)
  )

# integrity check: confirm every plot matched a treatment and water level (no rows silently dropped)
cat("\n=== Data Check ===\n")
print(paste("Total rows:", nrow(nmds_scores_yr0)))
print(paste("Missing trt values:", sum(is.na(nmds_scores_yr0$trt))))
print(paste("Missing water values:", sum(is.na(nmds_scores_yr0$water))))
print(head(nmds_scores_yr0))

nmds_scores_yr0 <- nmds_scores_yr0 %>% filter(!is.na(trt), !is.na(water))

# distance matrix subset matching the filtered scores (used for betadisper and pairwise PERMANOVA)
year_0bc_clean <- vegdist(as.matrix(year_0[rownames(nmds_scores_yr0), ]), method = "bray")

adonis_trt_0 <- adonis2(year_0bc ~ trt, data = nmds_scores_yr0, permutations = 999)
print(adonis_trt_0)
adonis_h2o_0 <- adonis2(year_0bc ~ water, data = nmds_scores_yr0, permutations = 999)
print(adonis_h2o_0)

# --- PAIRWISE PERMANOVA: YEAR 0 ---
cat("\n=== Year 0 Pairwise PERMANOVA: Treatment ===\n")
pw_adonis_trt_0 <- pairwise.adonis(year_0bc_clean, nmds_scores_yr0$trt,
                                   sim.method = "bray", p.adjust.m = "BH", perm = 999)
print(pw_adonis_trt_0)

cat("\n=== Year 0 Pairwise PERMANOVA: Precipitation ===\n")
pw_adonis_h2o_0 <- pairwise.adonis(year_0bc_clean, nmds_scores_yr0$water,
                                   sim.method = "bray", p.adjust.m = "BH", perm = 999)
print(pw_adonis_h2o_0)
```

## Betadisper: Year 0

```{r betadisper-year-0}
# --- BETADISPER: YEAR 0 ---
disp_trt_0 <- betadisper(year_0bc_clean, nmds_scores_yr0$trt)

disp_h2o_0 <- betadisper(year_0bc_clean, nmds_scores_yr0$water)
```

## Figure 3A, 2019 NMDS by seed mix

```{r plot-5-treatment-year0-14}
p_trt_yr0 <- gg_ordiplot(ca.nmds_yr0, groups = nmds_scores_yr0$trt, ellipse = TRUE, plot = FALSE)$plot +
  labs(color = "Seed Mixes", x = "NMDS1", y = "NMDS2",
       title = paste0("2019 by Seed Mix (Stress = ", round(ca.nmds_yr0$stress, 3), ")")) +
  scale_color_viridis_d(option = "viridis", begin = 0.2, end = 0.9) +
  annotate("text", x = -Inf, y = Inf,
           hjust = -0.05,
           vjust = 1.2,
           size = 6,
           lineheight = 0.9,
           label = paste0("PERMANOVA: R² = ", round(adonis_trt_0$R2[1], 3),
                          "\n p = ", round(adonis_trt_0$`Pr(>F)`[1], 3)),
           family = "Arial") +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  coord_fixed() +
  theme_nmds() +
  theme(aspect.ratio = 1,
        panel.grid   = element_blank(),
        plot.title   = element_text(size = 18, face = "bold", family = "Arial"),
        axis.title   = element_text(size = 16, face = "bold", family = "Arial"),
        axis.text    = element_text(size = 14, family = "Arial"),
        legend.title = element_text(size = 16, face = "bold", family = "Arial"),
        legend.text  = element_text(size = 16, family = "Arial"),
        legend.key.width = unit(0.4, "cm"))

print(p_trt_yr0)
ggsave("A.5.18.26.png", plot = p_trt_yr0, width = 8, height = 6, dpi = 1200, units = "in", bg = "white")
```

## Figure 3C, 2019 NMDS by precipitation

```{r plot-6-water-year0-15}
p_water_yr0 <- gg_ordiplot(ca.nmds_yr0, groups = nmds_scores_yr0$water, ellipse = TRUE, plot = FALSE)$plot +
  labs(color = "Precipitation", x = "NMDS1", y = "NMDS2",
       title = paste0("2019 by Precipitation (Stress = ", round(ca.nmds_yr0$stress, 3), ")")) +
  scale_color_viridis_d(option = "viridis", begin = 0.2, end = 0.6, labels = c("Addition", "Reduction")) +
  annotate("text", x = -Inf, y = Inf,
           hjust = -0.05,
           vjust = 1.2,
           size = 6,
           lineheight = 0.9,
           label = paste0("PERMANOVA: R² = ", round(adonis_h2o_0$R2[1], 3),
                          "\n p = ", round(adonis_h2o_0$`Pr(>F)`[1], 3)),
           family = "Arial") +
  guides(color = guide_legend(override.aes = list(size = 3))) +
  coord_fixed() +
  theme_nmds() +
  theme(aspect.ratio = 1,
        panel.grid   = element_blank(),
        plot.title   = element_text(size = 18, face = "bold", family = "Arial"),
        axis.title   = element_text(size = 16, face = "bold", family = "Arial"),
        axis.text    = element_text(size = 14, family = "Arial"),
        legend.title = element_text(size = 16, face = "bold", family = "Arial"),
        legend.text  = element_text(size = 16, family = "Arial"),
        legend.key.width = unit(0.4, "cm"))
print(p_water_yr0)
ggsave("B.5.18.26.png", plot = p_water_yr0, width = 8, height = 6, dpi = 1200, units = "in", bg = "white")
```

## analysis 6: NMDS year 4

```{r analysis-6-nmds-year-4-16}
year_4[is.na(year_4)] <- 0
year_4bc <- vegdist(as.matrix(year_4), method = "bray")
ca.nmds_yr4 <- metaMDS(year_4bc, k = 3, trymax = 100, trace = FALSE)

nmds_scores_yr4 <- as.data.frame(scores(ca.nmds_yr4, display = "sites"))

rownames_yr4 <- rownames(nmds_scores_yr4)

nmds_scores_yr4 <- nmds_scores_yr4 %>%
  mutate(
    rowname_original = rownames_yr4,
    trt = tolower(sub(" .*", "", rowname_original)),
    water = as.character(water.levels[rowname_original, "water"])
  ) %>%
  mutate(
    trt = factor(trt, levels = c("r", "dt", "fd", "ir")),
    trt = recode_factor(trt,
                        "r" = "Random",
                        "dt" = "Drought Tolerant",
                        "fd" = "Functionally Diverse",
                        "ir" = "Invasion Resistant"),
    water = factor(water, levels = water_levels, labels = water_labels)
  )

# integrity check: confirm every plot matched a treatment and water level (no rows silently dropped)
cat("\n=== Year 4 Data Check ===\n")
print(paste("Total rows:", nrow(nmds_scores_yr4)))
print(paste("Missing trt values:", sum(is.na(nmds_scores_yr4$trt))))
print(paste("Missing water values:", sum(is.na(nmds_scores_yr4$water))))
print(head(nmds_scores_yr4))

nmds_scores_yr4_clean <- nmds_scores_yr4 %>% filter(!is.na(trt), !is.na(water))

# Distance matrix subset matching the filtered scores (used for betadisper and pairwise PERMANOVA)
year_4bc_clean <- vegdist(as.matrix(year_4[rownames(nmds_scores_yr4_clean), ]), method = "bray")

adonis_trt_4 <- adonis2(year_4bc ~ trt, data = nmds_scores_yr4_clean, permutations = 999)
adonis_h2o_4 <- adonis2(year_4bc ~ water, data = nmds_scores_yr4_clean, permutations = 999)

# --- PAIRWISE PERMANOVA: YEAR 4 ---
cat("\n=== Year 4 Pairwise PERMANOVA: Treatment ===\n")
pw_adonis_trt_4 <- pairwise.adonis(year_4bc_clean, nmds_scores_yr4_clean$trt,
                                   sim.method = "bray", p.adjust.m = "BH", perm = 999)
print(pw_adonis_trt_4)

cat("\n=== Year 4 Pairwise PERMANOVA: Precipitation ===\n")
pw_adonis_h2o_4 <- pairwise.adonis(year_4bc_clean, nmds_scores_yr4_clean$water,
                                   sim.method = "bray", p.adjust.m = "BH", perm = 999)
print(pw_adonis_h2o_4)
```

## Betadisper: Year 4

```{r betadisper-year-4}
# --- BETADISPER: YEAR 4 ---
disp_trt_4 <- betadisper(year_4bc_clean, nmds_scores_yr4_clean$trt)

disp_h2o_4 <- betadisper(year_4bc_clean, nmds_scores_yr4_clean$water)
```

### Figure 3B, 2023 NMDS by seed mix

```{r plot-7-treatment-year4-19}
  p_trt_yr4 <- gg_ordiplot(ca.nmds_yr4, groups = nmds_scores_yr4_clean$trt, ellipse = TRUE, plot = FALSE)$plot +
  labs(color = "Seed Mixes", x = "NMDS1", y = "NMDS2",
       title = paste0("2023 by Seed Mix (Stress = ", round(ca.nmds_yr4$stress, 3), ")")) +
  scale_color_viridis_d(option = "viridis", begin = 0.2, end = 0.9) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1,
           vjust = 1.2, size = 6, lineheight = 0.9,
           label = paste0("PERMANOVA: R² = ", round(adonis_trt_4$R2[1], 3),
                          "\n p = ", round(adonis_trt_4$`Pr(>F)`[1], 3)),
           family = "Arial")+
  guides(color = guide_legend(override.aes = list(size = 3))) +
  coord_fixed() +
  theme_nmds() +
  theme(aspect.ratio = 1,
        panel.grid   = element_blank(),
        plot.title   = element_text(size = 18, face = "bold", family = "Arial"),
        axis.title   = element_text(size = 16, face = "bold", family = "Arial"),
        axis.text    = element_text(size = 14, family = "Arial"),
        legend.title = element_text(size = 16, face = "bold", family = "Arial"),
        legend.text  = element_text(size = 16, family = "Arial"),
        legend.key.width = unit(0.4, "cm"))
print(p_trt_yr4)
ggsave("C.5.18.26.png", plot = p_trt_yr4, width = 8, height = 6, dpi = 1200, units = "in", bg = "white")
```

### Figure 3D, 2023 NMDS by precipitation

```{r plot-8-water-year4-20}
  p_water_yr4 <- gg_ordiplot(ca.nmds_yr4, groups = nmds_scores_yr4_clean$water, ellipse = TRUE, plot = FALSE)$plot +
  labs(color = "Precipitation", x = "NMDS1", y = "NMDS2",
       title = paste0("2023 by Precipitation (Stress = ", round(ca.nmds_yr4$stress, 3), ")")) +
  scale_color_viridis_d(option = "viridis", begin = 0.2, end = 0.6, labels = c("Addition", "Reduction")) +
  annotate("text", x = Inf, y = Inf, hjust = 1.1,
           vjust = 1.2, size = 6, lineheight = 0.9,
           label = paste0("PERMANOVA: R² = ", round(adonis_h2o_4$R2[1], 3),
                          "\n p = ", round(adonis_h2o_4$`Pr(>F)`[1], 3)),
           family = "Arial")+
  guides(color = guide_legend(override.aes = list(size = 3))) +
  coord_fixed() +
  theme_nmds() +
  theme(aspect.ratio = 1,
        panel.grid   = element_blank(),
        plot.title   = element_text(size = 18, face = "bold", family = "Arial"),
        axis.title   = element_text(size = 16, face = "bold", family = "Arial"),
        axis.text    = element_text(size = 14, family = "Arial"),
        legend.title = element_text(size = 16, face = "bold", family = "Arial"),
        legend.text  = element_text(size = 16, family = "Arial"),
        legend.key.width = unit(0.4, "cm"))

print(p_water_yr4)
ggsave("D.5.18.26.png", plot = p_water_yr4, width = 8, height = 6, dpi = 1200, units = "in", bg = "white")

# summary print out of permanovas, betadisper, pairwise permanova's 
# --- CONSOLE SUMMARY PRINTOUTS ---
cat("\n=== Year 0 Treatment PERMANOVA ===\n"); print(adonis_trt_0)
cat("\n=== Year 0 Water PERMANOVA ===\n"); print(adonis_h2o_0)
cat("\n=== Year 4 Treatment PERMANOVA ===\n"); print(adonis_trt_4)
cat("\n=== Year 4 Water PERMANOVA ===\n"); print(adonis_h2o_4)

cat("\n=== Year 0 Treatment BETADISPER ANOVA ===\n"); print(anova(disp_trt_0))
cat("\n=== Year 0 Water BETADISPER ANOVA ===\n"); print(anova(disp_h2o_0))
cat("\n=== Year 4 Treatment BETADISPER ANOVA ===\n"); print(anova(disp_trt_4))
cat("\n=== Year 4 Water BETADISPER ANOVA ===\n"); print(anova(disp_h2o_4))

cat("\n=== Year 0 Pairwise PERMANOVA: Treatment ===\n"); print(pw_adonis_trt_0)
cat("\n=== Year 0 Pairwise PERMANOVA: Precipitation ===\n"); print(pw_adonis_h2o_0)
cat("\n=== Year 4 Pairwise PERMANOVA: Treatment ===\n"); print(pw_adonis_trt_4)
cat("\n=== Year 4 Pairwise PERMANOVA: Precipitation ===\n"); print(pw_adonis_h2o_4)
print(betadisper(year_4bc_clean, nmds_scores_yr4_clean$water)$group.distances)
```

## Figure 3: NMDS ordinations (4 panels)

```{r figure-3-nmds-4-panel}
# Pull out each shared legend (left-justified) so both rows can reserve the SAME
# width for the plot panels -> all four ordination boxes end up identical in size.
leg_seedmix <- get_legend(p_trt_yr0   + theme(legend.justification = "left"))  # shared by A & B
leg_precip  <- get_legend(p_water_yr0 + theme(legend.justification = "left"))  # shared by C & D

# Top row: seed-mix ordinations (A = 2019, B = 2023) + shared legend on the right.
row_AB <- ggarrange(
  p_trt_yr0 + theme(legend.position = "none"),
  p_trt_yr4 + theme(legend.position = "none"),
  as_ggplot(leg_seedmix),
  labels     = c("A", "B", ""),
  label.y = 0.95,
  ncol = 3, nrow = 1, widths = c(1, 1, 0.5),
  font.label = list(size = 22, face = "bold", family = "Arial")
)

# Bottom row: precipitation ordinations (C = 2019, D = 2023) + shared legend on the right.
row_CD <- ggarrange(
  p_water_yr0 + theme(legend.position = "none"),
  p_water_yr4 + theme(legend.position = "none"),
  as_ggplot(leg_precip),
  labels     = c("C", "D", ""),
  label.y = 0.95,
  ncol = 3, nrow = 1, widths = c(1, 1, 0.5),
  font.label = list(size = 22, face = "bold", family = "Arial")
)

# Stack the two rows; identical width splits keep all four boxes the same size.
figure_3 <- ggarrange(row_AB, row_CD, ncol = 1, nrow = 2)
print(figure_3)
ggsave("Figure_3.png", plot = figure_3, width = 14, height = 13, dpi = 1200, units = "in", bg = "white")
```

## analysis 7: shannon diversity and richness above and belowground

```{r analysis-7-shannon-diversity-and-richness-above-and-belowground-21}
# 1. Clean and match aboveground columns
abvg_clean <- abvg_4 %>% dplyr::select(-1); rownames(abvg_clean) <- abvg_4[, 1]
abvg_clean <- abvg_clean[, !colnames(abvg_clean) %in% c("BRDI", "BRHO", "BRMA", "FEMY", "FEPE", "HOMU", "POMO")]
colnames(abvg_clean)[colnames(abvg_clean) == "LUMII"] <- "LUMI"
colnames(abvg_clean)[colnames(abvg_clean) == "CACI"] <- "CAME"
abvg_matched <- abvg_clean[, match(colnames(native.rel.rev), colnames(abvg_clean))]

# drop empty rows to avoid NaN math warnings
abvg_matched_clean <- abvg_matched[rowSums(abvg_matched, na.rm = TRUE) > 0, ]

# 2. to handle spaces, dashes, or underscores
calc_indices_robust <- function(matrix_data, type_label, water_df) {
  
  df_metrics <- data.frame(
    shannon = diversity(matrix_data, index = "shannon"),
    richness = specnumber(matrix_data)
  ) %>%
    rownames_to_column(var = "rowname") %>%
    mutate(rowname_clean = gsub("[-_]", " ", rowname)) %>%
    separate(rowname_clean, into = c("part1", "part2"), sep = "\\s+", fill = "right", remove = TRUE) %>%
    mutate(
      part1_upper = toupper(part1),
      part2_upper = toupper(part2),
      treatment = case_when(
        part1_upper %in% c("R", "DT", "FD", "IR") ~ part1_upper,
        part2_upper %in% c("R", "DT", "FD", "IR") ~ part2_upper,
        TRUE ~ NA_character_
      ),
      plot = case_when(
        grepl("^[0-9]+$", part1) ~ part1,
        grepl("^[0-9]+$", part2) ~ part2,
        TRUE ~ NA_character_
      ),
      type = type_label
    ) %>%
    dplyr::select(rowname, shannon, richness, treatment, plot, type)
  
  water_df_clean <- water_df %>%
    mutate(
      treatment = toupper(as.character(treatment)),
      plot = as.character(plot)
    )
  
  df_metrics %>%
    left_join(water_df_clean, by = c("treatment", "plot"))
}

# 3. re-run index extractions
df_above <- calc_indices_robust(decostand(abvg_matched_clean, "total"), "above", abvg_water)
df_below <- calc_indices_robust(native.rel.rev, "below", abvg_water)

# 4. make combined dataset and clean missing groups
above_below <- bind_rows(df_above, df_below) %>%
  mutate(
    treatment = factor(treatment, levels = trt_levels),
    type = factor(type, levels = c("above", "below")),
    water = factor(water, levels = water_levels)
  ) %>%
  filter(!is.na(treatment), !is.na(water), !is.na(structure), !is.na(shannon))

# 5. Three-Way Mixed-Effects Models (Type III ANOVA with Satterthwaite's method via lmerTest)
shannon_model_full <- lmer(shannon ~ treatment * type * water + (1|structure), data = above_below)
summary(shannon_model_full)
anova(shannon_model_full)

richness_model_full <- lmer(richness ~ treatment * type * water + (1|structure), data = above_below)
summary(richness_model_full)
print(anova(richness_model_full))

letters_shannon <- get_cld_letters(shannon_model_full, above_below, "shannon")
letters_richness <- get_cld_letters(richness_model_full, above_below, "richness")
```

## Figure S4, above/belowground Shannon diversity

```{r plot-9-above-and-below-shannon-boxplot-22}
above_below_shannon_boxplot <- ggplot(above_below, aes(y = shannon, x = treatment, fill = interaction(type, water))) +
  geom_boxplot(outlier.shape = NA, position = position_dodge(width = 0.75)) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "darkred",
               position = position_dodge(width = 0.75), aes(group = interaction(type, water))) +
  geom_text(data = letters_shannon, aes(y = yposition, label = Letters, group = interaction(type, water)),
            position = position_dodge(width = 0.75), vjust = 0, size = 6.5, show.legend = FALSE, family = "Arial") +
  scale_x_discrete(labels = trt_labels) +
  scale_fill_viridis_d(option = "viridis", begin = 0.2, end = 0.9, name = "",
                       labels = c("above.125" = "Aboveground - Precip Addition", "above.50" = "Aboveground - Precip Reduction",
                                  "below.125" = "Belowground - Precip Addition", "below.50" = "Belowground - Precip Reduction")) +
  scale_y_continuous(expand = expansion(mult = c(0.07, 0.15))) +
  labs(x = NULL, y = "Native Shannon Diversity") +
  theme_custom() +
  theme(
    panel.grid      = element_blank(),                                  # remove background grid
    axis.line       = element_line(colour = "black", linewidth = 0.8),  # solid axis lines
    axis.ticks      = element_line(colour = "black", linewidth = 0.8),
    legend.text     = element_text(size = 20, family = "Arial"),
    legend.title    = element_text(size = 20, family = "Arial"),
    legend.key.size = unit(1.2, "cm"),
    axis.title.y    = element_text(size = 20, face = "bold", family = "Arial"),
    axis.text.x     = element_text(size = 20, family = "Arial"),
    axis.text.y     = element_text(size = 20, family = "Arial")
  ) +
  # stack the 4 legend entries into two rows at the bottom
  guides(fill = guide_legend(nrow = 2, byrow = TRUE))

print(above_below_shannon_boxplot)
ggsave("Figure_S4.png", plot = above_below_shannon_boxplot, width = 12, height = 6, dpi = 1200, units = "in", bg = "white")
```

## Figure S3, above/belowground richness

```{r plot-10-above-and-below-richness-boxplot-23}
above_below_richness_boxplot <- ggplot(above_below, aes(y = richness, x = treatment, fill = interaction(type, water))) +
  geom_boxplot(outlier.shape = NA, position = position_dodge(width = 0.75)) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, color = "darkred",
               position = position_dodge(width = 0.75), aes(group = interaction(type, water))) +
  geom_text(data = letters_richness, aes(y = yposition, label = Letters, group = interaction(type, water)),
            position = position_dodge(width = 0.75), vjust = 0, size = 6.5, show.legend = FALSE, family = "Arial") +
  scale_x_discrete(labels = trt_labels) +
  scale_fill_viridis_d(option = "viridis", begin = 0.2, end = 0.9, name = "",
                       labels = c("above.125" = "Aboveground - Precip Addition", "above.50" = "Aboveground - Precip Reduction",
                                  "below.125" = "Belowground - Precip Addition", "below.50" = "Belowground - Precip Reduction")) +
  scale_y_continuous(expand = expansion(mult = c(0.07, 0.15))) +
  labs(x = NULL, y = "Native Species Richness") +
  theme_custom() +
  theme(
    panel.grid      = element_blank(),                                  # remove background grid
    axis.line       = element_line(colour = "black", linewidth = 0.8),  # solid axis lines
    axis.ticks      = element_line(colour = "black", linewidth = 0.8),
    legend.text     = element_text(size = 20, family = "Arial"),
    legend.title    = element_text(size = 20, family = "Arial"),
    legend.key.size = unit(1.2, "cm"),
    axis.title.y    = element_text(size = 20, face = "bold", family = "Arial"),
    axis.text.x     = element_text(size = 20, family = "Arial"),
    axis.text.y     = element_text(size = 20, family = "Arial")
  ) +
  # stack the 4 legend entries into two rows at the bottom
  guides(fill = guide_legend(nrow = 2, byrow = TRUE))

print(above_below_richness_boxplot)
ggsave("Figure_S3.png", plot = above_below_richness_boxplot, width = 12, height = 6, dpi = 1200, units = "in", bg = "white")
```

## analysis 8: species abundance comparisons

```{r analysis-8-species-abundance-comparisons-24}
cat("\n=== Year 0 Data ===\n")
print(head(year_0))
print(paste("Dimensions:", nrow(year_0), "x", ncol(year_0)))

cat("\n=== Year 4 Data ===\n")
print(head(year_4))
print(paste("Dimensions:", nrow(year_4), "x", ncol(year_4)))

# calculate means before converting to full names
species_means_yr0 <- data.frame(
  code = names(colMeans(year_0, na.rm = TRUE)),
  mean_abundance = colMeans(year_0, na.rm = TRUE),
  row.names = NULL
)

species_means_yr4 <- data.frame(
  code = names(colMeans(year_4, na.rm = TRUE)),
  mean_abundance = colMeans(year_4, na.rm = TRUE),
  row.names = NULL
)

cat("\n=== Species Means Year 0 ===\n")
print(head(species_means_yr0))

cat("\n=== Species Means Year 4 ===\n")
print(head(species_means_yr4))

# create a species lookup table
species_lookup <- data.frame(
  code = c("FEMI", "PHCI", "BRCA", "AMME", "ACAM", "GRCA", "CAME", "CIOC",
           "ACST", "SPAI", "KOMA", "ACMI", "ESCA", "LUSU", "MURI", "ELGL",
           "SIBE", "LUBI", "DEFA", "MEIM", "LUMI", "HOIN", "STPU", "ASFA",
           "POSE", "ARPU", "LAPL", "RACA", "PLER", "TRWI", "HOBR", "ELCO"),
  species_name = c(
    "F. microstachys", "P. cicutaria", "B. carinatus", "A. menziesii",
    "A. americanus", "G. camporum", "C. menziesii", "C. occidentale",
    "A. strigosus", "S. airoides", "K. macrantha", "A. millefolium",
    "E. californica", "L. succulentus", "M. rigens", "E. glaucus",
    "S. bellum", "L. bicolor", "D. fasciculata", "M. imperfecta",
    "L. microcarpus", "H. intercedens", "S. pulchra", "A. fascicularis",
    "P. secunda", "A. purpurea", "L. platyglossa", "R. californicus",
    "P. erecta", "T. willdenovii", "H. brachyantherum", "E. condensatus"
  ),
  stringsAsFactors = FALSE
)

# join with lookup table
species_means_yr0 <- species_means_yr0 %>%
  left_join(species_lookup, by = "code") %>%
  dplyr::select(code, species_name, mean_abundance) %>%
  mutate(species_name = ifelse(is.na(species_name), code, species_name)) %>%
  dplyr::select(species_name, mean_abundance) %>%
  rename(species = species_name)

species_means_yr4 <- species_means_yr4 %>%
  left_join(species_lookup, by = "code") %>%
  dplyr::select(code, species_name, mean_abundance) %>%
  mutate(species_name = ifelse(is.na(species_name), code, species_name)) %>%
  dplyr::select(species_name, mean_abundance) %>%
  rename(species = species_name)

cat("\n=== After Name Conversion - Yr0 ===\n")
print(head(species_means_yr0))

cat("\n=== After Name Conversion - Yr4 ===\n")
print(head(species_means_yr4))

# define order
species_order_years <- bind_rows(species_means_yr0, species_means_yr4) %>%
  group_by(species) %>%
  summarise(overall_mean = mean(mean_abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(overall_mean)) %>%
  pull(species)

species_means_yr0$species <- factor(species_means_yr0$species, levels = species_order_years)
species_means_yr4$species <- factor(species_means_yr4$species, levels = species_order_years)

y_range_years <- range(c(species_means_yr0$mean_abundance, species_means_yr4$mean_abundance), na.rm = TRUE)
```

## Figure 4, species means 2019 (top panel)

```{r plot-11-species-means-year-0-25}
p_year0_bars <- ggplot(species_means_yr0, aes(x = species, y = mean_abundance, fill = species)) +
  geom_col(show.legend = FALSE) +
  ylim(y_range_years) +
  labs(x = NULL, y = NULL) +
  scale_fill_manual(values = rep(viridisLite::viridis(1, begin = 0.2), length(unique(species_means_yr0$species)))) +
  scale_x_discrete(labels = setNames(species_lookup$code, species_lookup$species_name)) +  # show 4-letter codes
  annotate("text", x = Inf, y = Inf, label = "2019", hjust = 1.3, vjust = 1.8,
           size = 7, fontface = "bold", family = "Arial") +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    panel.grid  = element_blank(),                                  # remove gridlines
    axis.line   = element_line(colour = "black", linewidth = 0.6),  # solid axis lines
    axis.ticks  = element_line(colour = "black", linewidth = 0.6),
    axis.text.x = element_text(angle = 55, vjust = 1, hjust = 1, size = 14, face = "bold", family = "Arial"),
    axis.text.y = element_text(size = 14, family = "Arial"),
    plot.title  = element_text(size = 16, face = "bold", family = "Arial", hjust = 0.5),
    plot.margin = margin(t = 5, r = 5, b = 22, l = 5)  # extra bottom room so angled codes aren't clipped
  )

print(p_year0_bars)
```

## Figure 4, species means 2023 (bottom panel + combine)

```{r plot-12-species-means-year-4-26}
p_year4_bars <- ggplot(species_means_yr4, aes(x = species, y = mean_abundance, fill = species)) +
  geom_col(show.legend = FALSE) +
  ylim(y_range_years) +
  labs(x = NULL, y = NULL) +
  scale_fill_manual(values = rep(viridisLite::viridis(1, begin = 0.2), length(unique(species_means_yr0$species)))) +
  scale_x_discrete(labels = setNames(species_lookup$code, species_lookup$species_name)) +  # show 4-letter codes
  annotate("text", x = Inf, y = Inf, label = "2023", hjust = 1.3, vjust = 1.8,
           size = 7, fontface = "bold", family = "Arial") +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    panel.grid  = element_blank(),                                  # remove gridlines
    axis.line   = element_line(colour = "black", linewidth = 0.6),  # solid axis lines
    axis.ticks  = element_line(colour = "black", linewidth = 0.6),
    axis.text.x = element_text(angle = 55, vjust = 1, hjust = 1, size = 14, face = "bold", family = "Arial"),
    axis.text.y = element_text(size = 14, family = "Arial"),
    plot.title  = element_text(size = 16, face = "bold", family = "Arial", hjust = 0.5),
    plot.margin = margin(t = 5, r = 5, b = 22, l = 5)  # extra bottom room so angled codes aren't clipped
  )

print(p_year4_bars)

# Top 10 species by year
top_yr0 <- species_means_yr0 %>% arrange(desc(mean_abundance)) %>% head(10)
top_yr4 <- species_means_yr4 %>% arrange(desc(mean_abundance)) %>% head(10)

print(top_yr0)
print(top_yr4)

year_comparison_figure <- suppressWarnings(ggarrange(p_year0_bars, p_year4_bars,
                                                     ncol = 1, nrow = 2)) %>%
  annotate_figure(bottom = text_grob("Species", size = 18, face = "bold", family = "Arial"),
                  left = text_grob("Relative Abundance", size = 18, face = "bold", family = "Arial", rot = 90))
print(year_comparison_figure)

ggsave("Figure_4.png", plot = year_comparison_figure, width = 12, height = 10, dpi = 1200, units = "in", bg = "white")

## above & belowground comparison figure:
abvg_matched_clean <- abvg_matched[rowSums(abvg_matched, na.rm = TRUE) > 0, ]
abvg_4_rel <- decostand(abvg_matched_clean, method = "total")

species_means_abvg <- data.frame(
  code = names(colMeans(abvg_4_rel, na.rm = TRUE)),
  mean_abundance = colMeans(abvg_4_rel, na.rm = TRUE),
  row.names = NULL
) %>%
  left_join(species_lookup, by = "code") %>%
  mutate(species_name = ifelse(is.na(species_name), code, species_name)) %>%
  dplyr::select(species_name, mean_abundance) %>%
  rename(species = species_name)

species_means_blwg <- data.frame(
  code = names(colMeans(year_4, na.rm = TRUE)),
  mean_abundance = colMeans(year_4, na.rm = TRUE),
  row.names = NULL
) %>%
  left_join(species_lookup, by = "code") %>%
  mutate(species_name = ifelse(is.na(species_name), code, species_name)) %>%
  dplyr::select(species_name, mean_abundance) %>%
  rename(species = species_name)

# find shared species
shared_species <- intersect(
  as.character(species_means_abvg$species),
  as.character(species_means_blwg$species)
)

# determine global ordering based on shared species
species_order_loc <- bind_rows(species_means_abvg, species_means_blwg) %>%
  filter(species %in% shared_species) %>%
  group_by(species) %>%
  summarise(overall_mean = mean(mean_abundance, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(overall_mean)) %>%
  pull(species)

# filter datasets down to shared species before factoring
species_means_abvg <- species_means_abvg %>% filter(species %in% shared_species)
species_means_blwg <- species_means_blwg %>% filter(species %in% shared_species)

# apply factors 
species_means_abvg$species <- factor(species_means_abvg$species, levels = species_order_loc)
species_means_blwg$species <- factor(species_means_blwg$species, levels = species_order_loc)

y_range_loc <- range(c(species_means_abvg$mean_abundance, species_means_blwg$mean_abundance), na.rm = TRUE)
```

## Figure S5, species means aboveground (top panel)

```{r plot-13-species-means-aboveground-27}
p_above_bars <- ggplot(species_means_abvg, aes(x = species, y = mean_abundance, fill = species)) +
  geom_col(show.legend = FALSE) +
  ylim(y_range_loc) +
  labs(x = NULL, y = NULL) +
  scale_fill_manual(values = rep(viridisLite::viridis(1, begin = 0.2), length(unique(species_means_abvg$species)))) +
  scale_x_discrete(labels = setNames(species_lookup$code, species_lookup$species_name)) +  # show 4-letter codes
  annotate("text", x = Inf, y = Inf, label = "Aboveground", hjust = 1.1, vjust = 1.8,
           size = 7, fontface = "bold", family = "Arial") +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    panel.grid  = element_blank(),                                  # remove gridlines
    axis.line   = element_line(colour = "black", linewidth = 0.6),  # solid axis lines
    axis.ticks  = element_line(colour = "black", linewidth = 0.6),
    axis.text.x = element_text(angle = 55, vjust = 1, hjust = 1, size = 14, face = "bold", family = "Arial"),
    axis.text.y = element_text(size = 14, family = "Arial"),
    plot.title  = element_text(size = 16, face = "bold", family = "Arial", hjust = 0.5),
    plot.margin = margin(t = 5, r = 5, b = 22, l = 5)  # extra bottom room so angled codes aren't clipped
  )
```

## Figure S5, species means belowground (bottom panel + combine)

```{r plot-14-species-means-belowground-28}
p_below_bars <- ggplot(species_means_blwg, aes(x = species, y = mean_abundance, fill = species)) +
  geom_col(show.legend = FALSE) +
  ylim(y_range_loc) +
  labs(x = NULL, y = NULL) +
  scale_fill_manual(values = rep(viridisLite::viridis(1, begin = 0.2), length(unique(species_means_blwg$species)))) +
  scale_x_discrete(labels = setNames(species_lookup$code, species_lookup$species_name)) +  # show 4-letter codes
  annotate("text", x = Inf, y = Inf, label = "Belowground", hjust = 1.1, vjust = 1.8,
           size = 7, fontface = "bold", family = "Arial") +
  coord_cartesian(clip = "off") +
  theme_minimal() +
  theme(
    panel.grid  = element_blank(),                                  # remove gridlines
    axis.line   = element_line(colour = "black", linewidth = 0.6),  # solid axis lines
    axis.ticks  = element_line(colour = "black", linewidth = 0.6),
    axis.text.x = element_text(angle = 55, vjust = 1, hjust = 1, size = 14, face = "bold", family = "Arial"),
    axis.text.y = element_text(size = 14, family = "Arial"),
    plot.title  = element_text(size = 16, face = "bold", family = "Arial", hjust = 0.5),
    plot.margin = margin(t = 5, r = 5, b = 22, l = 5)  # extra bottom room so angled codes aren't clipped
  )

location_comparison_figure <- ggarrange(p_above_bars, p_below_bars,
                                        ncol = 1, nrow = 2) %>%
  annotate_figure(bottom = text_grob("Species", size = 18, face = "bold", family = "Arial"),
                  left = text_grob("Relative Abundance", size = 18, face = "bold", family = "Arial", rot = 90))

print(location_comparison_figure)
ggsave("Figure_S5.png", plot = location_comparison_figure, width = 12, height = 10, dpi = 1200, units = "in", bg = "white")
```

### analysis 9: coupling matrix analysis (permanova and betadisper)

```{r analysis-9-coupling-matrix-analysis-permanova-and-betadisper-29}
abvg_4_rel <- decostand(abvg_matched_clean, method = "total")
year_4_rel <- decostand(native_mat.rev7, method = "total")

# function to standardize row names so they match perfectly
standardize_rownames <- function(mat) {
  rn <- rownames(mat)
  rn <- tolower(gsub("[-_]", " ", rn))
  rn <- gsub("^([0-9]+)\\s+([a-z]+)$", "\\2 \\1", rn)
  rownames(mat) <- rn
  return(mat)
}

# apply the fix to both matrices
abvg_4_rel <- standardize_rownames(abvg_4_rel)
year_4_rel <- standardize_rownames(year_4_rel)

# row matching fix
common_plots <- intersect(rownames(abvg_4_rel), rownames(year_4_rel))

abvg_4_rel <- abvg_4_rel[common_plots, ]
year_4_rel <- year_4_rel[common_plots, ]

allcomms9 <- bind_rows(
  as.data.frame(abvg_4_rel),
  as.data.frame(year_4_rel)
)
allcomms9[is.na(allcomms9)] <- 0

allcomms9$dummy_spp <- 0.0001

n_plots <- nrow(abvg_4_rel)
rownames(allcomms9) <- c(
  paste0(rownames(abvg_4_rel), "_above"),
  paste0(rownames(year_4_rel), "_below")
)

bcdist_comm_raw9 <- vegdist(as.matrix(allcomms9), method = "bray")
bcdist_matrix_raw9 <- as.matrix(bcdist_comm_raw9)

bcdist_comm_subset <- bcdist_matrix_raw9[1:n_plots, (n_plots + 1):(2 * n_plots)]
coupling_dissimilarity <- data.frame(
  id = rownames(abvg_4_rel),
  dist.ca = diag(bcdist_comm_subset)
)

coupling_dissimilarity$trt <- sub(" .*", "", coupling_dissimilarity$id)

coupling_dissimilarity <- coupling_dissimilarity %>%
  separate(id, into = c("treatment", "plot_num"), sep = " ", remove = FALSE) %>%
  mutate(treatment = toupper(treatment)) %>%
  left_join(abvg_water %>% mutate(plot = as.character(plot), treatment = toupper(treatment)),
            by = c("treatment", "plot_num" = "plot")) %>%
  dplyr::select(id, dist.ca, trt = treatment, water)

set.seed(123)
ca.nmds_coupling <- metaMDS(bcdist_comm_raw9, distance = "bray", k = 2, trymax = 100)

nmds_scores_coupling <- as.data.frame(scores(ca.nmds_coupling, display = "sites"))
nmds_scores_coupling$type <- factor(rep(c("Above", "Below"), each = n_plots),
                                    levels = c("Above", "Below"))

cat("\n=== PERMANOVA: COMMUNITY TYPE ===\n")
adonis_type <- adonis2(bcdist_comm_raw9 ~ type, data = nmds_scores_coupling, permutations = 999)
print(adonis_type)

cat("\n=== Pairwise PERMANOVA: Community Type (Above vs Below) ===\n")
pw_adonis_type <- pairwise.adonis(bcdist_comm_raw9, nmds_scores_coupling$type,
                                  sim.method = "bray", p.adjust.m = "BH", perm = 999)
print(pw_adonis_type)

cat("\n=== BETADISPER: COMMUNITY TYPE ===\n")
disper_type <- betadisper(bcdist_comm_raw9, group = nmds_scores_coupling$type)
print(disper_type)

cat("\n--- ANOVA of Betadisper (Are dispersions significantly different?) ---\n")
print(anova(disper_type))

# extracting values for plot annotation
r2_type <- round(adonis_type$R2[1], 3)
p_type  <- adonis_type$`Pr(>F)`[1]
```

## Figure 5, above- vs belowground NMDS

```{r plot-15-above-and-below-nmds-plot-30}
coupling_plot <- gg_ordiplot(ca.nmds_coupling, groups = nmds_scores_coupling$type,
                             ellipse = TRUE, plot = FALSE)

final_coupling_plot <- coupling_plot$plot +
  labs(
    color = "Community Type",
    x = "NMDS1",
    y = "NMDS2",
    title = paste0("2023 Above vs Belowground (Stress = ", round(ca.nmds_coupling$stress, 3), ")")
  ) +
  scale_color_viridis_d(option = "viridis", begin = 0.2, end = 0.6,
                        labels = c("Aboveground", "Belowground")) +
  annotate("text", x = -Inf, y = Inf,
           label = paste0("PERMANOVA:\nR² = ", r2_type, ", p = ", p_type),
           hjust = -0.1, vjust = 1.2, size = 4.5, family = "Arial") +
  theme_nmds() +
  theme(
    panel.grid   = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.8),
    plot.margin  = margin(t = 10, r = 10, b = 35, l = 35),
    plot.title   = element_text(size = 14, face = "bold", family = "Arial"),
    axis.title   = element_text(size = 16, face = "bold", family = "Arial"),
    axis.text    = element_text(size = 14, family = "Arial"),
    legend.title = element_text(size = 16, face = "bold", family = "Arial"),
    legend.text  = element_text(size = 14, family = "Arial"),
    legend.key.width = unit(0.4, "cm")
  ) +
  coord_cartesian(clip = "off")

print(final_coupling_plot)
ggsave("Figure_5.png", plot = final_coupling_plot, width = 8, height = 6, dpi = 1200, units = "in", bg = "white")
```

### analysis 10: contrasts

```{r analysis-10-contrasts-31}
invasive <- apply_factors(invasive)
model_fp <- glmer.nb(total_invasive ~ treatment * water + (1|structure), data = invasive)

emm_fp <- emmeans(model_fp, ~ water | treatment)
contrasts_fp <- contrast(emm_fp, method = "revpairwise", by = "treatment")
contrast_summary <- summary(contrasts_fp, infer = TRUE)
```

### Figure S2, forest plot (precipitation effect)

```{r plot-16-forest-plot-32, fig.width = 9, fig.height = 4}
  forest_data <- as.data.frame(contrast_summary) %>%
  mutate(
    ci95_lower = asymp.LCL,
    ci95_upper = asymp.UCL,
    ci50_lower = estimate - 0.674 * SE,
    ci50_upper = estimate + 0.674 * SE,
    treatment = factor(treatment, levels = trt_levels)
  )

forest_plot <- ggplot(forest_data, aes(x = estimate, y = treatment, color = treatment)) +
  geom_errorbarh(aes(xmin = ci95_lower, xmax = ci95_upper), height = 0, linewidth = 0.5) +
  geom_errorbarh(aes(xmin = ci50_lower, xmax = ci50_upper), height = 0, linewidth = 1.5) +
  geom_point(size = 4) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_viridis_d(option = "viridis", begin = 0.2, end = 0.9) +
  scale_y_discrete(labels = trt_labels) +
  labs(x = "Effect of precipitation on invasive seed counts\n(Addition - Reduction)", y = NULL) +
  theme_classic(base_family = "Arial") +
  theme(
    axis.text.y = element_text(size = 16, family = "Arial"),
    axis.text.x = element_text(size = 14, family = "Arial"),
    axis.title  = element_text(size = 18, face = "bold", family = "Arial"),
    legend.position = "none"
  )

print(forest_plot)
ggsave("Figure_S2.png", plot = forest_plot, width = 9, height = 4, dpi = 1200, units = "in", bg = "white")
```

## analysis 11: seed mass correlation

```{r analysis-11-seed-mass-correlation-33}
abundance_mass <- lm(mean_abundance ~ seed.mass, data = seed_data)
print(summary(abundance_mass))
print(anova(abundance_mass))
```

